---
title: "Bayesian Hierarchical Drug-Response Model: Gibbs Sampler"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This notebook implements and validates the Gibbs sampler for the
network-structured horseshoe drug-response model (Section 4 of
`pathway.tex`), in four parts:

1. **Sampler functions** -- the model implementation itself (every full
   conditional is Normal or Inverse-Gamma; there is no
   Metropolis-Hastings step anywhere).
2. **Simulation-based calibration** -- validates the sampler on data with
   a known true answer, *before* trusting it on real data. This step
   previously caught a genuine model-degeneracy bug (see the writeup in
   Section 6 below); it is kept in this notebook rather than removed, so
   the same check reruns automatically if you ever modify the model.
3. **Real-data run** -- runs the validated sampler on your actual
   `mutation_matrix.csv` / `response_matrix.csv` /
   `pathway_adjacency_W.csv` files, using **multiple independent chains**
   so the diagnostics in Part 4 (which require $\geq 2$ chains, e.g.
   Gelman-Rubin) can be computed.
4. **MCMC diagnostics** -- Gelman-Rubin $\hat R$ (across chains),
   effective sample size (ESS), and the Geweke diagnostic (within each
   chain), with an automatic flagging table so you don't have to eyeball
   hundreds of parameters by hand. Calibration on simulated data
   (Part 2) checks whether the *posterior* is right given enough samples;
   these diagnostics check whether the *chain itself* is behaving like a
   converged, well-mixing sample in the first place -- a different
   question, not previously checked in this notebook, and one that
   matters here in particular because of a known correlated gene cluster
   in the real data (see the project progress log's multicollinearity
   diagnostic) that could plausibly slow mixing for the genes involved.

Run the chunks in order: Part 1 defines functions (fast), Part 2 runs a
calibration check (a few minutes), Part 3 runs on real data (time depends
on your iteration settings -- a short test setting is provided first),
Part 4 runs diagnostics on whatever Part 3 produced.

---

# Part 1: Sampler Functions

```{r libraries}
suppressWarnings(suppressMessages({
  library(Matrix)
}))

# coda provides the standard implementations of Gelman-Rubin (gelman.diag),
# effective sample size (effectiveSize), and the Geweke diagnostic
# (geweke.diag) used in Part 4 -- these are not reimplemented here, since
# coda's versions are the standard, widely-checked reference
# implementations in R.
if (!requireNamespace("coda", quietly = TRUE)) {
  install.packages("coda", repos = "https://cloud.r-project.org")
}
suppressWarnings(suppressMessages({
  library(coda)
}))
```

```{r rinvgamma}
# R has no built-in Inverse-Gamma sampler without an extra package;
# implemented directly from rgamma to avoid an additional dependency,
# since IG(a, b) <=> 1 / Gamma(a, rate = b).
rinvgamma <- function(n, shape, rate) {
  1 / rgamma(n, shape = shape, rate = rate)
}
```


```{r gibbs-sampler-function}
gibbs_sampler <- function(y, M, group, W, delta = 0.5,
                           use_group_layer = TRUE,
                           use_network_layer = TRUE,
                           use_horseshoe_layer = TRUE,
                           flat_prior_var = 100,
                           n_iter = 4000, burn_in = 2000, thin = 2,
                           verbose = 200, seed = 1,
                           gigg_a = 1, gigg_b = 1) {
  # gigg_a, gigg_b: shape/rate hyperparameters for the Gamma prior on the
  # precision of each GLOBAL shrinkage parameter (1/kappa2_d, 1/zeta2_d).
  # Gamma(1,1) is the default used during validation on real data (see
  # project progress log); both parameters stabilized at sensible,
  # non-degenerate values under this choice across all genes and drugs
  # tested. As with delta, a small sensitivity grid over (gigg_a, gigg_b)
  # is recommended before treating any single choice as final for the
  # paper's results.
  #
  # use_network_layer, use_horseshoe_layer: set either to FALSE to fit the
  # corresponding ABLATION comparator specified in pathway.tex Phase 3
  # (Section 3.4): a no-network variant (genes exchangeable, closer to
  # Samorodnitsky et al. 2020's pooling structure) or a no-horseshoe
  # variant (flat priors on gene effects, no sparsity layer). Both
  # default to TRUE (the full model); setting both FALSE simultaneously
  # is allowed and fits beta_tilde with ONLY the flat prior below -- a
  # useful third comparator isolating the data term alone. When a layer
  # is disabled, its associated parameters (kappa2/W/delta for the
  # network layer; lambda2/zeta2/nu for the horseshoe layer) are not
  # estimated and are excluded from the returned output for that fit,
  # rather than being computed and silently ignored.
  #
  # flat_prior_var: the prior variance used for beta_tilde when
  # use_horseshoe_layer = FALSE, i.e. beta_tilde_gd ~ Normal(0,
  # flat_prior_var) independently per gene. Default 100, matching the
  # flat prior already used for alpha_kd elsewhere in this model -- a
  # genuinely uninformative prior across mutation-effect coefficients'
  # realistic range (which is empirically O(0.01) to O(1), per the real
  # gene/coefficient values observed during validation), not a
  # tightly-tuned value chosen to make this comparator look artificially
  # good or bad.

  set.seed(seed)

  N <- nrow(y)
  D <- ncol(y)
  G <- ncol(M)
  gene_names <- colnames(M)

  stopifnot(nrow(M) == N)
  stopifnot(length(group) == N)

  K <- length(unique(group))
  group_idx <- as.integer(factor(group))  # remap to 1..K contiguous

  obs_mask <- !is.na(y)  # N x D logical: TRUE where y is observed

  # --- Network prior setup ---
  # Genes in M that also appear in W get the GMRF network prior on
  # beta_tilde; genes in M absent from W (e.g. the 83 unmapped genes in
  # the real panel) get horseshoe-only shrinkage, handled by giving them
  # an "infinite-variance" (i.e. zero-precision contribution) row/column
  # in the network precision matrix below rather than excluding them from
  # the model entirely. If use_network_layer = FALSE (the no-network
  # ablation comparator, pathway.tex Phase 3), G_net is forced to 0
  # regardless of W, which disables the network term entirely through the
  # same `if (G_net > 0)` gate already used for genes absent from W --
  # no separate code path is needed for this ablation.
  if (use_network_layer) {
    net_genes <- intersect(gene_names, rownames(W))
  } else {
    net_genes <- character(0)
  }
  net_idx <- match(net_genes, gene_names)         # positions within 1..G
  G_net <- length(net_genes)

  if (G_net > 0) {
    W_sub <- W[net_genes, net_genes, drop = FALSE]
    D_W <- diag(rowSums(W_sub))
    Prec_net_sub <- D_W - delta * W_sub  # (D_W - delta*W), to be scaled by 1/kappa_d^2
    # Verify positive-definiteness once, up front, rather than discovering
    # a failure deep inside the sampler loop -- this mirrors the check
    # already performed during network construction (see the project
    # progress log) and is repeated here defensively in case W or delta
    # changes between sessions.
    eigvals_check <- eigen(Prec_net_sub, symmetric = TRUE, only.values = TRUE)$values
    if (min(eigvals_check) <= 1e-8) {
      stop(sprintf(
        "D_W - delta*W is not positive definite at delta=%.3f (min eigenvalue = %.6e). Re-run the network construction step or choose a different delta from the verified grid {0.5, 0.7, 0.8, 0.9}.",
        delta, min(eigvals_check)
      ))
    }
  } else {
    Prec_net_sub <- matrix(numeric(0), 0, 0)
  }

  # --- Initialize parameters ---
  alpha <- matrix(0, K, D)
  beta_tilde <- matrix(0, G, D)
  beta <- array(0, dim = c(K, G, D))  # group-level beta (used only if use_group_layer)
  sigma2 <- rep(1, D)
  eta2 <- matrix(1, G, D)
  kappa2 <- rep(1, D)
  lambda2 <- matrix(1, G, D)
  zeta2 <- rep(1, D)
  nu <- matrix(1, G, D)    # horseshoe LOCAL auxiliary variable (lambda2 stays Inverse-Gamma)
  # Note: the horseshoe GLOBAL auxiliary variable "xi" used in the original
  # Makalic-Schmidt half-Cauchy-via-IG-mixture scheme is no longer needed --
  # zeta2's precision is now drawn directly from a Gamma distribution (see
  # Step 7 below and the file header), which is already a complete
  # conjugate family requiring no further auxiliary-variable augmentation.

  # --- Storage for kept samples ---
  keep_iters <- seq(burn_in + 1, n_iter, by = thin)
  n_kept <- length(keep_iters)
  out_alpha <- array(NA, dim = c(K, D, n_kept))
  out_beta_tilde <- array(NA, dim = c(G, D, n_kept))
  out_sigma2 <- matrix(NA, D, n_kept)
  out_kappa2 <- matrix(NA, D, n_kept)
  out_lambda2 <- array(NA, dim = c(G, D, n_kept))
  out_zeta2 <- matrix(NA, D, n_kept)
  keep_pos <- 0L

  # Precompute, per drug, which rows (samples) are observed -- avoids
  # repeating the is.na() scan every iteration.
  obs_rows_by_drug <- lapply(seq_len(D), function(d) which(obs_mask[, d]))

  # ----------------------------------------------------------------- #
  # PERFORMANCE: precompute every per-drug quantity that does NOT depend
  # on the current iteration's sampled parameters, once, before the
  # iteration loop -- rather than recomputing it fresh on every one of
  # n_iter iterations. The dominant cost identified by profiling on real
  # data (see the project progress log) was crossprod(M_d) in the
  # no-group-layer path, which is a function of the fixed mutation data
  # alone and is therefore identical on every iteration for a given drug;
  # recomputing a ~951x219 crossprod thousands of times per drug was
  # responsible for roughly 60% of total runtime. M_d itself (the row
  # slice of M for drug d's observed samples) is precomputed for the same
  # reason: slicing is cheap per call but still unnecessary repeated work
  # across thousands of iterations.
  # ----------------------------------------------------------------- #
  M_d_list <- vector("list", D)
  y_d_list <- vector("list", D)
  grp_d_list <- vector("list", D)
  data_prec_full_unscaled_list <- vector("list", D)  # crossprod(M_d), no-group-layer path only

  for (d in seq_len(D)) {
    rows_d <- obs_rows_by_drug[[d]]
    if (length(rows_d) == 0) next
    M_d_list[[d]] <- M[rows_d, , drop = FALSE]
    y_d_list[[d]] <- y[rows_d, d]
    grp_d_list[[d]] <- group_idx[rows_d]
    if (!use_group_layer) {
      data_prec_full_unscaled_list[[d]] <- crossprod(M_d_list[[d]])
    }
  }

  for (t in seq_len(n_iter)) {

    for (d in seq_len(D)) {
      rows_d <- obs_rows_by_drug[[d]]
      if (length(rows_d) == 0) next  # drug screened on zero retained samples; skip

      y_d <- y_d_list[[d]]
      M_d <- M_d_list[[d]]
      grp_d <- grp_d_list[[d]]

      # current fitted mutation contribution, used to back out the
      # intercept's residual and vice versa
      mu_mut_d <- M_d %*% beta_tilde[, d]  # using beta_tilde directly if no group layer;
                                            # overwritten below if group layer active
      if (use_group_layer) {
        beta_d_per_row <- beta[grp_d, , d, drop = FALSE]
        # recompute mu_mut_d using the group-specific beta for each row
        mu_mut_d <- rowSums(M_d * beta[grp_d, , d])
      }

      # ---------------------------------------------------------------- #
      # Step 1: intercepts alpha_{kd}
      # ---------------------------------------------------------------- #
      for (k in seq_len(K)) {
        rows_k <- rows_d[grp_d == k]
        if (length(rows_k) == 0) next
        idx_in_d <- which(grp_d == k)
        resid_k <- y_d[idx_in_d] - mu_mut_d[idx_in_d]
        prec_k <- length(rows_k) / sigma2[d] + 1 / 100
        mean_k <- (sum(resid_k) / sigma2[d]) / prec_k
        alpha[k, d] <- rnorm(1, mean_k, sqrt(1 / prec_k))
      }

      # recompute residual against updated intercepts
      resid_d <- y_d - alpha[grp_d, d]

      # ---------------------------------------------------------------- #
      # Step 2 & 3: group-level beta and group heterogeneity variance
      # (optional layer only)
      # ---------------------------------------------------------------- #
      if (use_group_layer) {
        for (k in seq_len(K)) {
          idx_in_d <- which(grp_d == k)
          if (length(idx_in_d) == 0) {
            beta[k, , d] <- rnorm(G, beta_tilde[, d], sqrt(eta2[, d]))
            next
          }
          M_k <- M_d[idx_in_d, , drop = FALSE]
          y_k <- resid_d[idx_in_d]
          # Conjugate Normal update for a G-vector regression coefficient
          # with independent Normal(beta_tilde, eta2) prior per gene.
          prior_prec <- diag(1 / eta2[, d], G, G)
          A_k <- crossprod(M_k) / sigma2[d] + prior_prec
          b_k <- crossprod(M_k, y_k) / sigma2[d] + (beta_tilde[, d] / eta2[, d])
          # Single Cholesky factorization serves both the posterior mean
          # and the random draw -- see the identical optimization and its
          # rationale documented above Step 4's B_d update.
          R_k <- chol(A_k)
          beta_mean <- backsolve(R_k, forwardsolve(t(R_k), b_k))
          beta[k, , d] <- as.numeric(
            beta_mean + backsolve(R_k, rnorm(G))
          )
        }
        # Step 3: group heterogeneity variance eta2_gd
        for (g in seq_len(G)) {
          ss <- sum((beta[, g, d] - beta_tilde[g, d])^2)
          eta2[g, d] <- rinvgamma(1, shape = K / 2 + 0.01, rate = 0.01 + ss / 2)
        }
      }

      # ---------------------------------------------------------------- #
      # Step 4: network-and-horseshoe-structured mean effects beta_tilde_.d
      # ---------------------------------------------------------------- #
      if (use_group_layer) {
        # Data-driven term comes from pooling across groups (sum over k of
        # beta_{kgd}), precision diag(1/eta2_gd) * K in the simplest
        # complete-pooling case; here we use the exact per-group precision
        # sum so unequal eta2_gd values, if they arise, are respected.
        data_prec_diag <- rep(0, G)
        data_b <- rep(0, G)
        for (k in seq_len(K)) {
          data_prec_diag <- data_prec_diag + 1 / eta2[, d]
          data_b <- data_b + beta[k, , d] / eta2[, d]
        }
      } else {
        # No-group variant: beta_tilde regresses directly on the data.
        # data_prec_full_unscaled_list[[d]] = crossprod(M_d), precomputed
        # once before the iteration loop since it does not depend on any
        # sampled parameter -- this was the single largest cost in
        # profiling (see comment above the precomputation block).
        data_prec_full <- data_prec_full_unscaled_list[[d]] / sigma2[d]
        data_b_full <- crossprod(M_d, resid_d) / sigma2[d]
      }

      # Horseshoe precision (diagonal): 1 / (lambda_gd^2 * zeta_d^2).
      # If use_horseshoe_layer = FALSE (the no-horseshoe ablation
      # comparator, pathway.tex Phase 3), beta_tilde instead gets a flat,
      # fixed Normal(0, flat_prior_var) prior per gene -- a genuinely
      # uninformative prior (no sparsity, no shrinkage), not a
      # disguised replacement for the horseshoe. lambda2/zeta2/nu are not
      # estimated in this case (Steps 6-7 are skipped below).
      if (use_horseshoe_layer) {
        hs_prec_diag <- 1 / (lambda2[, d] * zeta2[d])
      } else {
        hs_prec_diag <- rep(1 / flat_prior_var, G)
      }

      # Network precision (only on the G_net genes that map to a pathway):
      # full G x G precision matrix, zero outside the net_idx submatrix,
      # so genes with no pathway membership get zero contribution here
      # (they rely entirely on the horseshoe term above, as documented in
      # pathway.tex Section 4). If use_network_layer = FALSE (the
      # no-network ablation comparator), G_net is forced to 0 above, so
      # this matrix is always all-zero here and contributes nothing --
      # kappa2 is correspondingly not estimated (Step 5 is skipped below).
      net_prec_full <- matrix(0, G, G)
      if (G_net > 0) {
        net_prec_full[net_idx, net_idx] <- Prec_net_sub / kappa2[d]
      }

      if (use_group_layer) {
        B_d <- diag(data_prec_diag, G, G) + net_prec_full + diag(hs_prec_diag, G, G)
        b_d <- data_b
      } else {
        B_d <- data_prec_full + net_prec_full + diag(hs_prec_diag, G, G)
        b_d <- as.numeric(data_b_full)
      }

      # PERFORMANCE: B_d is symmetric positive-definite (it's a precision
      # matrix: sum of a data-precision term, the GMRF network precision,
      # and the horseshoe precision, each PSD or PD), so it admits a
      # single Cholesky factorization R (upper-triangular, B_d = R'R) that
      # serves BOTH purposes below -- solving for the posterior mean via
      # two triangular solves, and drawing the random Normal perturbation
      # via a single triangular solve against a standard Normal vector.
      # This replaces the previous solve(B_d) (a full LU-based matrix
      # inverse) followed by a SECOND, separate chol() of that inverse,
      # which was redundant: both decompositions describe the same
      # underlying covariance structure, so doing it once is mathematically
      # equivalent and meaningfully faster (this and the analogous Step 2
      # update were the dominant remaining cost after the crossprod fix
      # documented above, per profiling on real data).
      R_d <- chol(B_d)  # upper-triangular, B_d = t(R_d) %*% R_d
      beta_tilde_mean <- backsolve(R_d, forwardsolve(t(R_d), b_d))
      beta_tilde[, d] <- as.numeric(
        beta_tilde_mean + backsolve(R_d, rnorm(G))
      )

      # ---------------------------------------------------------------- #
      # Step 5: network smoothness variance kappa2_d
      #
      # GIGG-style fix (see file header): kappa2_d's precision
      # (1/kappa2_d) is drawn from a Gamma distribution rather than
      # kappa2_d itself from an Inverse-Gamma -- this breaks the
      # distributional symmetry with zeta2_d (Step 7) that was previously
      # causing both to collapse toward zero together. Derivation: with
      # beta_tilde_net | tau_kappa ~ N(0, Prec_net^{-1} / tau_kappa), the
      # likelihood contribution for tau_kappa = 1/kappa2_d is
      # Gamma(shape=G_net/2, rate=quad_form/2); combined with a
      # Gamma(gigg_a, gigg_b) prior, the conjugate posterior is
      # Gamma(gigg_a + G_net/2, gigg_b + quad_form/2). Verified numerically
      # against a directly grid-evaluated posterior before adoption.
      # ---------------------------------------------------------------- #
      if (G_net > 0) {
        bt_net <- beta_tilde[net_idx, d]
        quad_form <- as.numeric(t(bt_net) %*% Prec_net_sub %*% bt_net)
        tau_kappa <- rgamma(1, shape = gigg_a + G_net / 2, rate = gigg_b + quad_form / 2)
        kappa2[d] <- 1 / tau_kappa
      }

      # ---------------------------------------------------------------- #
      # Step 6: horseshoe local shrinkage (per gene)
      #
      # Unchanged: lambda_gd keeps its original Inverse-Gamma (half-Cauchy
      # via Makalic-Schmidt augmentation) structure. GIGG prescribes
      # leaving the LOCAL shrinkage parameter Inverse-Gamma (heavy-tailed,
      # lets individual genes escape shrinkage when supported by data) and
      # only changing the GLOBAL/GROUP-level parameter's family -- see
      # Step 7 below and the file header. Skipped entirely if
      # use_horseshoe_layer = FALSE (the no-horseshoe ablation comparator,
      # pathway.tex Phase 3): lambda2/nu are not estimated in that case,
      # since beta_tilde instead uses the flat prior set up in Step 4.
      # ---------------------------------------------------------------- #
      if (use_horseshoe_layer) {
        for (g in seq_len(G)) {
          lambda2[g, d] <- rinvgamma(1, shape = 1,
                                      rate = 1 / nu[g, d] + beta_tilde[g, d]^2 / (2 * zeta2[d]))
          nu[g, d] <- rinvgamma(1, shape = 1, rate = 1 + 1 / lambda2[g, d])
        }
      }

      # ---------------------------------------------------------------- #
      # Step 7: horseshoe global shrinkage (per drug)
      #
      # GIGG-style fix (see file header and Step 5 above, applied
      # identically here): zeta2_d's precision is drawn from a Gamma
      # distribution instead of zeta2_d itself from an Inverse-Gamma.
      # zeta2_d has shape=(G+1)/2 (G=219 in the real panel, i.e. shape~110)
      # -- an even more severe instance of the Step 5 pathology, confirmed
      # on real data: zeta2_d collapsed to exactly 0 even after kappa2_d
      # was stabilized by the Step 5 fix alone, until this identical fix
      # was applied here too. The same derivation as Step 5 applies with
      # sum_term in place of quad_form and G in place of G_net. Skipped
      # entirely if use_horseshoe_layer = FALSE, same rationale as Step 6.
      # ---------------------------------------------------------------- #
      if (use_horseshoe_layer) {
        sum_term <- sum(beta_tilde[, d]^2 / lambda2[, d])
        tau_zeta <- rgamma(1, shape = gigg_a + G / 2, rate = gigg_b + sum_term / 2)
        zeta2[d] <- 1 / tau_zeta
      }

      # ---------------------------------------------------------------- #
      # Step 8: residual variance sigma2_d
      # ---------------------------------------------------------------- #
      if (use_group_layer) {
        mu_final <- alpha[grp_d, d] + rowSums(M_d * beta[grp_d, , d])
      } else {
        mu_final <- alpha[grp_d, d] + as.numeric(M_d %*% beta_tilde[, d])
      }
      ss_resid <- sum((y_d - mu_final)^2)
      sigma2[d] <- rinvgamma(1, shape = 0.01 + length(rows_d) / 2,
                              rate = 0.01 + ss_resid / 2)
    }

    if (verbose > 0 && t %% verbose == 0) {
      cat(sprintf("[gibbs_sampler] iteration %d / %d\n", t, n_iter))
    }

    if (t %in% keep_iters) {
      keep_pos <- keep_pos + 1L
      out_alpha[, , keep_pos] <- alpha
      out_beta_tilde[, , keep_pos] <- beta_tilde
      out_sigma2[, keep_pos] <- sigma2
      # Store NA (rather than the unused initialization value) for any
      # layer disabled via use_network_layer / use_horseshoe_layer, so a
      # saved ablation fit cannot be misread as having an active network
      # or horseshoe layer at some particular (meaningless) value -- the
      # config list above already records which layers were active, but
      # this makes the parameter arrays themselves self-consistent too.
      out_kappa2[, keep_pos] <- if (use_network_layer) kappa2 else NA
      out_lambda2[, , keep_pos] <- if (use_horseshoe_layer) lambda2 else NA
      out_zeta2[, keep_pos] <- if (use_horseshoe_layer) zeta2 else NA
    }
  }

  list(
    alpha = out_alpha,
    beta_tilde = out_beta_tilde,
    sigma2 = out_sigma2,
    kappa2 = out_kappa2,
    lambda2 = out_lambda2,
    zeta2 = out_zeta2,
    gene_names = gene_names,
    group_levels = levels(factor(group)),
    keep_iters = keep_iters,
    # Layer-usage metadata, so a saved fit is self-documenting about which
    # ablation configuration (if any) produced it -- relevant when
    # comparing multiple fits later via cross-validated predictive
    # likelihood (pathway.tex Phase 3).
    config = list(
      use_group_layer = use_group_layer,
      use_network_layer = use_network_layer,
      use_horseshoe_layer = use_horseshoe_layer,
      delta = delta,
      gigg_a = gigg_a,
      gigg_b = gigg_b,
      flat_prior_var = flat_prior_var
    )
  )
}
```

---

# Part 2: Simulation-Based Calibration

**Why this step exists, and why it is not optional:** before trusting any
result the sampler produces on real data, we check it against data where
the true answer is known. This previously caught a genuine bug: with the
optional group-pooling layer enabled and too few groups, the sampler
converges to a self-consistent but *wrong* answer (every network-linked
gene's effect collapses to zero regardless of its true value), driven by a
feedback loop between the group-heterogeneity variance and the
network-smoothness variance. It is **not** a coding error -- each
individual update was verified correct in isolation -- it is a real
property of the model when the number of groups is small relative to the
number of genes. The fix is not a code change but a **configuration**
choice: don't use the group-pooling layer unless you have enough groups
(rule of thumb from the diagnosis: tens of samples per group is fine; 4
groups for 15+ genes is not). The real dataset's 28 tissue types are
enough for the layer to behave correctly if you want to use it, but it is
not used by default below, matching the recommendation already in
`04_model.tex`.

```{r sim-data-function}
if (!requireNamespace("MASS", quietly = TRUE)) {
  install.packages("MASS", repos = "https://cloud.r-project.org")
}
library(MASS)

simulate_one_dataset <- function(N = 200, G = 15, D = 3, K = 4,
                                  G_net = 8, true_delta = 0.8) {
  W <- matrix(0, G_net, G_net)
  W[1:4, 1:4] <- 0.5
  W[5:8, 5:8] <- 0.5
  diag(W) <- 0
  W[4, 5] <- W[5, 4] <- 0.3
  rownames(W) <- colnames(W) <- paste0("gene", 1:G_net)
  gene_names <- paste0("gene", 1:G)

  true_alpha <- matrix(rnorm(K * D, 0, 1), K, D)
  true_kappa2 <- rep(0.5, D)
  true_sigma2 <- rep(0.3, D)

  D_W <- diag(rowSums(W))
  Prec_net <- D_W - true_delta * W
  Sigma_net <- solve(Prec_net) * true_kappa2[1]

  true_beta_tilde <- matrix(0, G, D)
  for (d in seq_len(D)) {
    true_beta_tilde[1:G_net, d] <- MASS::mvrnorm(1, rep(0, G_net), Sigma_net * true_kappa2[d] / true_kappa2[1])
    nonzero_extra <- sample((G_net + 1):G, 2)
    true_beta_tilde[nonzero_extra, d] <- rnorm(2, 0, 1.5)
  }

  mut_freq <- runif(G, 0.05, 0.4)
  M <- matrix(rbinom(N * G, 1, rep(mut_freq, each = N)), N, G)
  colnames(M) <- gene_names
  group <- sample(1:K, N, replace = TRUE)

  y <- matrix(NA, N, D)
  for (d in seq_len(D)) {
    mu <- true_alpha[group, d] + as.numeric(M %*% true_beta_tilde[, d])
    y[, d] <- rnorm(N, mu, sqrt(true_sigma2[d]))
  }

  list(y = y, M = M, group = group, W = W, true_alpha = true_alpha,
       true_beta_tilde = true_beta_tilde, true_sigma2 = true_sigma2,
       true_kappa2 = true_kappa2, gene_names = gene_names)
}
```

```{r calibration-function}
run_calibration <- function(use_group_layer, K, N, label, n_reps = 15,
                             n_iter = 1500, burn_in = 750, thin = 2) {
  cat(sprintf("\n=== Scenario: %s (K=%d groups, N=%d samples, use_group_layer=%s) ===\n",
              label, K, N, use_group_layer))

  coverage_beta_tilde <- c()
  coverage_sigma2 <- c()
  coverage_alpha <- c()

  for (rep_i in seq_len(n_reps)) {
    sim <- simulate_one_dataset(N = N, K = K)

    fit <- gibbs_sampler(
      y = sim$y, M = sim$M, group = sim$group, W = sim$W,
      delta = 0.8, use_group_layer = use_group_layer,
      n_iter = n_iter, burn_in = burn_in, thin = thin,
      verbose = 0, seed = 100 + rep_i
    )

    G <- nrow(sim$true_beta_tilde); D <- ncol(sim$true_beta_tilde)
    for (g in seq_len(G)) for (d in seq_len(D)) {
      ci <- quantile(fit$beta_tilde[g, d, ], c(0.025, 0.975))
      coverage_beta_tilde <- c(coverage_beta_tilde,
        sim$true_beta_tilde[g, d] >= ci[1] && sim$true_beta_tilde[g, d] <= ci[2])
    }
    for (d in seq_len(D)) {
      ci <- quantile(fit$sigma2[d, ], c(0.025, 0.975))
      coverage_sigma2 <- c(coverage_sigma2,
        sim$true_sigma2[d] >= ci[1] && sim$true_sigma2[d] <= ci[2])
    }
    Kk <- nrow(sim$true_alpha)
    for (k in seq_len(Kk)) for (d in seq_len(D)) {
      ci <- quantile(fit$alpha[k, d, ], c(0.025, 0.975))
      coverage_alpha <- c(coverage_alpha,
        sim$true_alpha[k, d] >= ci[1] && sim$true_alpha[k, d] <= ci[2])
    }
  }

  cat(sprintf("beta_tilde coverage: %.1f%% | sigma2: %.1f%% | alpha: %.1f%%\n",
              100 * mean(coverage_beta_tilde), 100 * mean(coverage_sigma2),
              100 * mean(coverage_alpha)))
  invisible(list(beta_tilde = mean(coverage_beta_tilde),
                  sigma2 = mean(coverage_sigma2), alpha = mean(coverage_alpha)))
}
```

```{r run-calibration}
# Set SIM_REPS lower (e.g. 5) for a quick smoke test, or higher (30+) for
# a more rigorous check. 15 is a reasonable default (a few minutes total).
n_reps <- 15

result_A <- run_calibration(use_group_layer = FALSE, K = 4, N = 200,
                             label = "(A) No-group-layer -- RECOMMENDED for real data",
                             n_reps = n_reps)
result_B <- run_calibration(use_group_layer = TRUE, K = 4, N = 200,
                             label = "(B) Group-layer, K=4 -- KNOWN FAILURE MODE (kept as a regression check)",
                             n_reps = n_reps)
result_C <- run_calibration(use_group_layer = TRUE, K = 20, N = 800,
                             label = "(C) Group-layer, K=20 -- viable group size",
                             n_reps = n_reps)
```

```{r calibration-summary}
summary_df <- data.frame(
  Scenario = c("(A) No-group (recommended)", "(B) Group layer, K=4 (failure mode)",
               "(C) Group layer, K=20 (viable)"),
  beta_tilde_coverage = sprintf("%.1f%%", 100 * c(result_A$beta_tilde, result_B$beta_tilde, result_C$beta_tilde)),
  sigma2_coverage = sprintf("%.1f%%", 100 * c(result_A$sigma2, result_B$sigma2, result_C$sigma2)),
  alpha_coverage = sprintf("%.1f%%", 100 * c(result_A$alpha, result_B$alpha, result_C$alpha))
)
summary_df
```

If (A) and (C) are both close to 95% and (B) is well below it (typically
60-70%), the sampler is behaving as expected and it is safe to proceed to
Part 3. If (A) is unexpectedly low, stop and investigate before trusting
any real-data result -- do not proceed to Part 3 in that case.

---

# Part 3: Run on Real Data

## Step 1: Load your real data files

Edit the paths below if your files aren't in the current working
directory.

**Important data note:** `gene_adjacency_W.csv` (a **gene x gene**
matrix, 135 genes after dropping one isolated node) is loaded here, NOT
`pathway_adjacency_W.csv` (a **pathway x pathway** matrix, 141 pathways).
An earlier version of this notebook loaded the pathway-level matrix by
mistake; since its row/column names are pathway names rather than gene
names, `intersect(gene_names, rownames(W))` inside `gibbs_sampler()`
silently evaluated to zero genes for every drug, which meant Step 5's
$\kappa_d^2$ update never ran at all and stayed frozen at its
initialization value (1.0) for the entire run -- this was confirmed
directly on real output (`kappa2` showing exactly 1.0 with zero variance
across thousands of kept samples, for every drug). `gene_adjacency_W.csv`
is derived from `pathway_gene_membership.csv` (two genes are linked if
they share at least one retained KEGG pathway), built and verified
positive-definite at the same candidate $\delta$ grid as the original
network; see the project progress log for the full construction and
diagnostic account.

```{r load-real-data}
mut <- read.csv("mutation_matrix.csv", row.names = 1, check.names = FALSE)
resp <- read.csv("response_matrix.csv")
groups <- read.csv("model_groups.csv")
W_real <- read.csv("gene_adjacency_W.csv", row.names = 1, check.names = FALSE)
W_real <- as.matrix(W_real)

cat("Mutation matrix:", nrow(mut), "genes x", ncol(mut), "models\n")
cat("Response rows:", nrow(resp), "| unique drugs:", length(unique(resp$drug_id)),
    "| unique models:", length(unique(resp$model_id)), "\n")
cat("Network matrix:", nrow(W_real), "x", ncol(W_real),
    "-- row/col names should be GENE symbols, not pathway names:\n")
print(head(rownames(W_real), 5))
```

## Step 2: Reshape into the sampler's expected format

```{r reshape-real-data}
model_ids <- colnames(mut)
M_real <- t(as.matrix(mut))          # N x G
rownames(M_real) <- model_ids

all_drugs <- sort(unique(resp$drug_id))
y_real <- matrix(NA, nrow = length(model_ids), ncol = length(all_drugs),
                  dimnames = list(model_ids, as.character(all_drugs)))
for (i in seq_len(nrow(resp))) {
  y_real[as.character(resp$model_id[i]), as.character(resp$drug_id[i])] <- resp$y[i]
}

# Group vector for the optional pooling layer -- NOT used by default
# (use_group_layer = FALSE below), per the recommendation in 04_model.tex
# and the calibration results in Part 2. Kept here in case you want to
# experiment with it.
group_vec <- groups$tissue[match(model_ids, groups$model_id)]

cat("y_real dim:", dim(y_real), "| NAs:", sum(is.na(y_real)), "/", length(y_real), "\n")
cat("Any model with no matching tissue group:", sum(is.na(group_vec)), "\n")
```

## Step 3: Short test run (recommended first)

This runs on the FULL real dataset (all 951 models, 219 genes) but only a
handful of drugs and a short chain, so it finishes in well under a minute
-- use this to confirm everything loads and runs correctly before
committing to a long full run. **Two chains** are run (not one), with
different seeds, so Part 4's diagnostics (which need $\geq 2$ chains for
Gelman-Rubin) can be exercised at this fast test scale before you commit
to a long full run.

```{r real-data-test-run}
test_drugs <- all_drugs[1:3]
y_test <- y_real[, as.character(test_drugs), drop = FALSE]

n_test_chains <- 2
fits_test <- vector("list", n_test_chains)

t0 <- Sys.time()
for (chain_i in seq_len(n_test_chains)) {
  fits_test[[chain_i]] <- gibbs_sampler(
    y = y_test, M = M_real, group = rep(1, length(model_ids)), W = W_real,
    delta = 0.8, use_group_layer = FALSE,
    n_iter = 200, burn_in = 100, thin = 2,
    verbose = 0, seed = chain_i
  )
}
t1 <- Sys.time()
cat("Elapsed:", round(as.numeric(difftime(t1, t0, units = "secs")), 1),
    "seconds for", length(test_drugs), "drugs,", n_test_chains,
    "chains, 200 iterations each.\n")
```

```{r real-data-test-sanity-check}
# Sanity check: TP53 is the most frequently mutated gene in the panel
# (67.3% of cell lines) and was the dominant predictor in an independent
# external study (Samorodnitsky et al. 2020, for survival rather than
# drug response) -- its posterior should show non-trivial signal, not
# collapse to a degenerate near-zero estimate the way the buggy
# group-layer configuration did in Part 2's Scenario B.
tp53_idx <- which(rownames(mut) == "TP53")
cat("TP53 beta_tilde posterior mean per test drug (chain 1):\n")
print(round(apply(fits_test[[1]]$beta_tilde[tp53_idx, , , drop = FALSE], c(1, 2), mean), 4))
cat("TP53 beta_tilde posterior SD per test drug (chain 1; should be > 0, not collapsed):\n")
print(round(apply(fits_test[[1]]$beta_tilde[tp53_idx, , , drop = FALSE], c(1, 2), sd), 4))
```

You can run Part 4's diagnostics on `fits_test` right now, before
committing to a full run, by setting `fits_for_diagnostics <- fits_test`
in the first chunk of Part 4 below -- useful for confirming the
diagnostic code itself works and for getting an early read on problem
genes at low cost, though short chains will generally show lower ESS and
higher $\hat R$ than a properly-sized full run, so don't treat test-scale
diagnostic results as a verdict on the model.

## Step 4: Full run (all drugs, multiple chains)

**This will take time, multiplied by the number of chains.** Edit the
settings in the chunk immediately below to change the chain length,
burn-in, thinning, or number of chains -- every other chunk in this
section reads from these variables, so you only need to change them in
one place.

```{r mcmc-settings}
# --- EDIT THESE TO CHANGE THE FULL RUN ---
mcmc_n_iter   <- 5000   # total iterations per chain
mcmc_burn_in  <- 1000   # discarded from the start of each chain
mcmc_thin     <- 2      # keep every mcmc_thin-th post-burn-in iteration
mcmc_n_chains <- 2      # >= 2 required for Gelman-Rubin in Part 4

# Samples kept per chain = (mcmc_n_iter - mcmc_burn_in) / mcmc_thin
samples_per_chain <- (mcmc_n_iter - mcmc_burn_in) / mcmc_thin
cat("Samples kept per chain:", samples_per_chain, "\n")
cat("Total chains:", mcmc_n_chains, "\n")
cat("Total kept samples across all chains:", samples_per_chain * mcmc_n_chains, "\n")
if (samples_per_chain != floor(samples_per_chain)) {
  warning("(mcmc_n_iter - mcmc_burn_in) is not evenly divisible by mcmc_thin -- the actual kept count will be floor()'d by the sampler; consider adjusting so the arithmetic is exact.")
}
```

With the defaults above (4{,}000 / 1{,}000 / 3), each chain keeps 1{,}000
samples. Runtime scales linearly in `mcmc_n_iter` and in `mcmc_n_chains`
(chains are run sequentially below, not in parallel); per-chain timing
benchmarked on this real dataset (951 models, 219 genes, all 295 drugs):

- `mcmc_n_iter = 1500`: ~45-50 minutes per chain
- `mcmc_n_iter = 2000`: ~65-70 minutes per chain
- `mcmc_n_iter = 4000`: ~2.2 hours per chain

These are real measured rates from this notebook's optimized sampler, not
rough guesses -- but your machine's speed will vary, and **the default
`mcmc_n_chains = 2` doubles whichever of the above applies.** At
`mcmc_n_iter = 4000` this means roughly 4.4 hours total for 2 chains.
Consider running this chunk separately (not as part of "Run All") and
saving the result with `saveRDS()` so you don't need to rerun it every
time you reopen this notebook.

```{r real-data-full-run, eval=FALSE}
# Set eval=TRUE above (or run this chunk manually) when you're ready for
# the full run. Left as eval=FALSE by default so "Run All" doesn't
# accidentally trigger a multi-hour run.

fits_full <- vector("list", mcmc_n_chains)

for (chain_i in seq_len(mcmc_n_chains)) {
  cat(sprintf("\n=== Chain %d / %d ===\n", chain_i, mcmc_n_chains))
  t0 <- Sys.time()
  fits_full[[chain_i]] <- gibbs_sampler(
    y = y_real, M = M_real, group = rep(1, length(model_ids)), W = W_real,
    delta = 0.5, use_group_layer = FALSE,
    n_iter = mcmc_n_iter, burn_in = mcmc_burn_in, thin = mcmc_thin,
    verbose = 200,
    seed = chain_i   # DIFFERENT SEED PER CHAIN -- required for chains to
                      # be independent draws from different starting
                      # points, which is what Gelman-Rubin in Part 4
                      # actually tests.
  )
  t1 <- Sys.time()
  cat(sprintf("Chain %d elapsed: %.1f minutes\n", chain_i,
              as.numeric(difftime(t1, t0, units = "mins"))))
}

saveRDS(fits_full, "gibbs_fits_full0.5.rds")
cat("\nSaved", mcmc_n_chains, "chains to gibbs_fits_full0.5.rds\n")
```

```{r reload-full-run, eval=FALSE}
# If you're returning to this notebook after the full run already
# completed and was saved, skip the chunk above and just reload:
fits_full <- readRDS("gibbs_fits_full.rds")
cat("Reloaded", length(fits_full), "chains.\n")
```

## Step 5: Inspect results (after the full run)

```{r inspect-results, eval=FALSE}
# Run this after Step 4 completes (or after reloading saved fits with
# fits_full <- readRDS("gibbs_fits_full.rds")).
#
# Uses chain 1 for this inspection; Part 4 below checks whether chain 1
# agrees with the other chains closely enough that using just one is
# actually justified (that is precisely what Gelman-Rubin measures) --
# check Part 4 before reporting any single-chain result like this one as
# a final answer.

# Posterior inclusion: which genes have a beta_tilde credible interval
# that excludes zero for a given drug, as a simple sparsity summary
# (a more formal posterior-inclusion-probability approach, following
# Pham et al. 2018 Section 4.7's BFDR thresholding, is a natural next
# step but not implemented in this notebook).
drug_to_inspect <- colnames(y_real)[1]
d_idx <- which(colnames(y_real) == drug_to_inspect)

ci_lo <- apply(fits_full[[1]]$beta_tilde[, d_idx, ], 1, quantile, 0.025)
ci_hi <- apply(fits_full[[1]]$beta_tilde[, d_idx, ], 1, quantile, 0.975)
excludes_zero <- (ci_lo > 0) | (ci_hi < 0)

genes_flagged <- rownames(mut)[excludes_zero]
cat("Drug", drug_to_inspect, "-- genes whose 95% CI excludes zero (chain 1):\n")
print(genes_flagged)
```

## Step 6: Ablation comparators (pathway.tex Phase 3)

Per the methodology document's Phase 3, the full model should be compared
against two reduced variants to demonstrate that the network layer and
the horseshoe layer are each "earning their complexity," not just adding
parameters for no benefit:

- **No-network**: `use_network_layer = FALSE` -- genes are exchangeable
  under the horseshoe alone (closer to Samorodnitsky et al. 2020's
  pooling structure, without the GMRF/pathway structure).
- **No-horseshoe**: `use_horseshoe_layer = FALSE` -- a flat
  $\text{Normal}(0,$ `flat_prior_var`$)$ prior replaces the sparsity
  layer; no shrinkage, no escape-from-shrinkage behavior.

Both reuse every other setting (drugs, chains, iterations) from
`mcmc_n_iter`/`mcmc_burn_in`/`mcmc_thin`/`mcmc_n_chains` set in Step 4
above, so the three fits are directly comparable.

```{r ablation-fits, eval=FALSE}
# Run this AFTER Step 4's full run (fits_full) has completed and been
# saved. Like Step 4, this is gated behind eval=FALSE since it runs the
# sampler two more times at full scale -- expect roughly the same
# per-chain runtime as Step 4 for each ablation variant.

fits_nonetwork <- vector("list", mcmc_n_chains)
fits_nohorseshoe <- vector("list", mcmc_n_chains)

for (chain_i in seq_len(mcmc_n_chains)) {
  cat(sprintf("\n=== No-network ablation, Chain %d / %d ===\n", chain_i, mcmc_n_chains))
  fits_nonetwork[[chain_i]] <- gibbs_sampler(
    y = y_real, M = M_real, group = rep(1, length(model_ids)), W = W_real,
    delta = 0.8, use_group_layer = FALSE, use_network_layer = FALSE,
    n_iter = mcmc_n_iter, burn_in = mcmc_burn_in, thin = mcmc_thin,
    verbose = 200, seed = chain_i
  )
}
saveRDS(fits_nonetwork, "gibbs_fits_nonetwork.rds")
cat("Saved no-network ablation fits.\n")

for (chain_i in seq_len(mcmc_n_chains)) {
  cat(sprintf("\n=== No-horseshoe ablation, Chain %d / %d ===\n", chain_i, mcmc_n_chains))
  fits_nohorseshoe[[chain_i]] <- gibbs_sampler(
    y = y_real, M = M_real, group = rep(1, length(model_ids)), W = W_real,
    delta = 0.8, use_group_layer = FALSE, use_horseshoe_layer = FALSE,
    n_iter = mcmc_n_iter, burn_in = mcmc_burn_in, thin = mcmc_thin,
    verbose = 200, seed = chain_i
  )
}
saveRDS(fits_nohorseshoe, "gibbs_fits_nohorseshoe.rds")
cat("Saved no-horseshoe ablation fits.\n")
```

```{r ablation-reload, eval=FALSE}
# Reload all three fits if returning to this notebook later:
fits_full <- readRDS("gibbs_fits_full.rds")
fits_nonetwork <- readRDS("gibbs_fits_nonetwork.rds")
fits_nohorseshoe <- readRDS("gibbs_fits_nohorseshoe.rds")
```

### Comparing models via held-out predictive likelihood

Following \citet{samorodnitsky2020}'s Section 2.4 (cited in
`pathway.tex`), the recommended comparison metric is out-of-sample
log-posterior predictive likelihood under $k$-fold cross-validation. The
function below computes a simpler in-sample version (posterior-mean
squared error and log-likelihood at the posterior mean) as a quick first
check; replacing this with genuine held-out folds is a natural
extension, not yet implemented here, since it requires re-fitting each
model $k$ times rather than once.

```{r ablation-comparison, eval=FALSE}
# Quick in-sample comparison (NOT a substitute for proper k-fold
# out-of-sample predictive likelihood, but a fast sanity check that the
# three models actually differ in fit quality before investing in the
# full cross-validation).
compare_insample_fit <- function(fit_chains, y, M, label) {
  # Use chain 1's posterior mean for mu, sigma2
  beta_tilde_mean <- apply(fit_chains[[1]]$beta_tilde, c(1, 2), mean)
  sigma2_mean <- apply(fit_chains[[1]]$sigma2, 1, mean)
  alpha_mean <- apply(fit_chains[[1]]$alpha, c(1, 2), mean)

  D <- ncol(y)
  total_loglik <- 0
  total_sq_err <- 0
  total_n <- 0
  for (d in seq_len(D)) {
    obs_rows <- which(!is.na(y[, d]))
    if (length(obs_rows) == 0) next
    mu_d <- alpha_mean[1, d] + as.numeric(M[obs_rows, , drop = FALSE] %*% beta_tilde_mean[, d])
    resid <- y[obs_rows, d] - mu_d
    total_sq_err <- total_sq_err + sum(resid^2)
    total_n <- total_n + length(obs_rows)
    total_loglik <- total_loglik + sum(dnorm(y[obs_rows, d], mu_d, sqrt(sigma2_mean[d]), log = TRUE))
  }
  cat(sprintf("%-20s: in-sample log-lik = %.1f | RMSE = %.4f | n = %d\n",
              label, total_loglik, sqrt(total_sq_err / total_n), total_n))
  invisible(list(loglik = total_loglik, rmse = sqrt(total_sq_err / total_n)))
}

cat("=== In-sample comparison (quick check; replace with k-fold for the paper) ===\n")
result_full <- compare_insample_fit(fits_full, y_real, M_real, "Full model")
result_nonet <- compare_insample_fit(fits_nonetwork, y_real, M_real, "No-network")
result_nohs <- compare_insample_fit(fits_nohorseshoe, y_real, M_real, "No-horseshoe")
```

---

# Part 4: MCMC Diagnostics

**Why this part exists.** Part 2's simulation-based calibration checks a
different question than the one here: calibration asks "if I have enough
posterior samples, are my credible intervals the right size and in the
right place?" -- it does not check whether the samples you actually drew
behave like a converged, well-mixing chain. A sampler can pass
calibration on a short, clean simulated problem and still mix poorly on
a larger, more correlated real dataset. This matters concretely here
because the multicollinearity diagnostic (project progress log) found a
biologically real co-mutation cluster (RPL22, ACVR2A, BAX, BMPR2, KMT2B,
STAT5B, linked to microsatellite instability) with a correlation matrix
condition number of 208 -- exactly the kind of structure that can slow
Gibbs sampler mixing for the genes involved, since the chain has to take
small steps through a narrow, correlated ridge in the posterior rather
than moving freely.

Three diagnostics are computed below, using the `coda` package's standard
reference implementations:

- **Gelman-Rubin $\hat R$**: compares within-chain to between-chain
  variance across $\geq 2$ chains started from different seeds. Values
  close to 1 indicate the chains agree; the conventional threshold for
  concern is $\hat R > 1.1$.
- **Effective Sample Size (ESS)**: how many independent draws your
  autocorrelated MCMC samples are actually worth. **You specified ESS
  $<$ 800 as not acceptable** for this project -- the summary table below
  flags every parameter below that threshold explicitly.
- **Geweke diagnostic**: within a single chain, compares the mean of an
  early window to the mean of a late window; the underlying Z-score is a
  standard-normal test statistic under the null of stationarity, reported
  here as a two-sided p-value (flagged at $p < 0.05$, equivalent to
  $|Z| > 1.96$) for consistency with how significance is reported
  elsewhere in this project. A small p-value suggests the chain had not
  yet stabilized over the window compared.

## Step 1: Choose which fit to diagnose

```{r diagnostics-choose-fit}
# Point this at whichever set of chains you want to check:
#   - fits_test   (from Part 3, Step 3 -- fast, useful for an early look)
#   - fits_full   (from Part 3, Step 4 -- the real, full-length run)
#
# EDIT THIS LINE:
fits_for_diagnostics <- fits_full # fits_test   # change to fits_full once you have it

cat("Diagnosing", length(fits_for_diagnostics), "chain(s).\n")
cat("Samples per chain:", dim(fits_for_diagnostics[[1]]$beta_tilde)[3], "\n")
cat("Drugs in this fit:", dim(fits_for_diagnostics[[1]]$beta_tilde)[2], "\n")
cat("Genes in this fit:", dim(fits_for_diagnostics[[1]]$beta_tilde)[1], "\n")

if (length(fits_for_diagnostics) < 2) {
  warning("Gelman-Rubin requires >= 2 chains. Re-run Part 3 with n_test_chains >= 2 (test scale) or mcmc_n_chains >= 2 (full scale) before computing Rhat.")
}
```

## Step 2: Diagnostic functions

```{r diagnostics-functions}
# Builds a coda mcmc.list for a single beta_tilde[gene, drug, ] parameter
# across all available chains. Returns NULL (rather than erroring) if any
# chain has zero variance for this parameter, since coda's gelman.diag()
# fails outright on a zero-variance chain and a single bad parameter
# should not stop a scan over hundreds of others.
build_mcmc_list_beta <- function(fits, gene_idx, drug_idx) {
  chains <- lapply(fits, function(f) {
    vals <- f$beta_tilde[gene_idx, drug_idx, ]
    if (sd(vals) < 1e-10) return(NULL)
    coda::mcmc(as.matrix(vals))
  })
  if (any(sapply(chains, is.null))) return(NULL)
  coda::mcmc.list(chains)
}

# Same idea, for a scalar-per-drug parameter like sigma2[d, ] or kappa2[d, ].
build_mcmc_list_scalar <- function(fits, extract_fn, drug_idx) {
  chains <- lapply(fits, function(f) {
    vals <- extract_fn(f, drug_idx)
    if (sd(vals) < 1e-10) return(NULL)
    coda::mcmc(as.matrix(vals))
  })
  if (any(sapply(chains, is.null))) return(NULL)
  coda::mcmc.list(chains)
}

# Scans every beta_tilde[gene, drug] parameter (or a specified subset of
# drugs, to keep runtime manageable when D is large) and returns a data
# frame with Rhat, ESS, and the Geweke diagnostic (chain 1) for each.
#
# Geweke's Z-score is already a standard-normal test statistic under the
# null of chain stationarity (Geweke, 1992), so the two-sided p-value is
# simply 2 * pnorm(-abs(Z)) -- this is computed and flagged on directly
# (geweke_p < 0.05) rather than thresholding the Z-score itself, since a
# p-value reads consistently alongside the rest of this project's
# credible-interval-based reporting. Note this is very close numerically
# to the previous |Z| > 2 rule (Z = 1.96 <=> p = 0.05 exactly; Z = 2 <=>
# p = 0.0455), so very few, if any, parameters change flag status as a
# result of this change -- the point of switching is interpretability and
# consistency with how significance is reported elsewhere, not a
# materially different cutoff.
scan_beta_tilde_diagnostics <- function(fits, drug_indices = NULL, ess_threshold = 800) {
  G <- dim(fits[[1]]$beta_tilde)[1]
  D_total <- dim(fits[[1]]$beta_tilde)[2]
  if (is.null(drug_indices)) drug_indices <- seq_len(D_total)
  gene_names <- fits[[1]]$gene_names

  results <- vector("list", G * length(drug_indices))
  idx <- 1L
  for (d in drug_indices) {
    for (g in seq_len(G)) {
      mcmc_obj <- build_mcmc_list_beta(fits, g, d)
      if (is.null(mcmc_obj)) {
        rhat <- NA; ess <- NA; geweke_z <- NA; geweke_p <- NA
      } else {
        rhat <- if (length(fits) >= 2) {
          tryCatch(coda::gelman.diag(mcmc_obj, autoburnin = FALSE)$psrf[1, 1],
                   error = function(e) NA)
        } else NA
        ess <- tryCatch(sum(coda::effectiveSize(mcmc_obj)), error = function(e) NA)
        geweke_z <- tryCatch(coda::geweke.diag(mcmc_obj[[1]])$z[[1]],
                              error = function(e) NA)
        geweke_p <- if (is.na(geweke_z)) NA else 2 * pnorm(-abs(geweke_z))
      }
      results[[idx]] <- data.frame(
        gene = gene_names[g], drug_col_idx = d,
        rhat = rhat, ess = ess, geweke_z = geweke_z, geweke_p = geweke_p
      )
      idx <- idx + 1L
    }
  }
  out <- do.call(rbind, results)
  out$rhat_flag <- !is.na(out$rhat) & out$rhat > 1.1
  out$ess_flag <- !is.na(out$ess) & out$ess < ess_threshold
  out$geweke_flag <- !is.na(out$geweke_p) & out$geweke_p < 0.05
  out
}
```

## Step 3: Run the scan

By default this scans every gene for the **first 3 drugs** in the fit, to
keep runtime reasonable for a quick check (each parameter requires
building an `mcmc.list` and running three diagnostics on it). Increase
`drugs_to_scan` once you've confirmed this runs correctly and want full
coverage -- with `fits_full` at full scale (219 genes $\times$ 295 drugs
$=$ 64{,}605 parameters) this should still take at most a few minutes,
since each individual diagnostic computation is cheap; it has not been
benchmarked at that full scale in this notebook.

```{r run-diagnostic-scan}
n_drugs_in_fit <- dim(fits_for_diagnostics[[1]]$beta_tilde)[2]
drugs_to_scan <- seq_len(n_drugs_in_fit) # seq_len(min(3, n_drugs_in_fit))  # EDIT: e.g. seq_len(n_drugs_in_fit) for all drugs

diag_results <- scan_beta_tilde_diagnostics(
  fits_for_diagnostics, drug_indices = drugs_to_scan, ess_threshold = 800
)

cat("Parameters scanned:", nrow(diag_results), "\n")
cat("Flagged for Rhat > 1.1:", sum(diag_results$rhat_flag, na.rm = TRUE), "\n")
cat("Flagged for ESS < 800:", sum(diag_results$ess_flag, na.rm = TRUE), "\n")
cat("Flagged for Geweke p-value < 0.05:", sum(diag_results$geweke_flag, na.rm = TRUE), "\n")
cat("\nESS distribution:\n")
print(summary(diag_results$ess))
```

## Step 4: Inspect flagged parameters

```{r inspect-flagged}
any_flag <- diag_results$rhat_flag | diag_results$ess_flag | diag_results$geweke_flag
flagged <- diag_results[any_flag & !is.na(any_flag), ]
flagged <- flagged[order(flagged$ess), ]  # worst ESS first

cat(nrow(flagged), "of", nrow(diag_results), "parameters flagged by at least one diagnostic.\n\n")
flagged
```

```{r check-msi-cluster, eval=TRUE}
# Specifically check the MSI co-mutation cluster genes identified in the
# multicollinearity diagnostic (project progress log) -- these are the
# genes most likely, on substantive grounds, to show slower mixing due to
# their mutual correlation.
msi_cluster_genes <- c("RPL22", "ACVR2A", "BAX", "BMPR2", "KMT2B", "STAT5B")
msi_results <- diag_results[diag_results$gene %in% msi_cluster_genes, ]
cat("MSI-cluster gene diagnostics (genes present in this fit's panel):\n")
print(msi_results)
```

## Step 5: Trace plots for a specific parameter (visual check)

Numbers above can hide patterns a trace plot makes obvious (slow drift,
one chain stuck far from the others, etc.). Plot any flagged parameter
here by gene name and drug column index.

```{r trace-plot, fig.width=9, fig.height=4}
gene_to_plot <- "TP53"      # EDIT: any gene name from rownames(mut)
drug_col_to_plot <- 1        # EDIT: column index into fits_for_diagnostics[[1]]$beta_tilde

g_idx <- which(fits_for_diagnostics[[1]]$gene_names == gene_to_plot)
mcmc_obj <- build_mcmc_list_beta(fits_for_diagnostics, g_idx, drug_col_to_plot)

if (is.null(mcmc_obj)) {
  cat("This parameter had zero variance in at least one chain -- nothing to plot.\n")
} else {
  plot(mcmc_obj, main = sprintf("%s, drug column %d", gene_to_plot, drug_col_to_plot))
}
```

## Interpreting the results

- **If very few parameters are flagged** (a small percentage, and not
  concentrated in any particular gene cluster): the chain length and
  thinning you chose are adequate; proceed to use the posterior samples
  for inference.
- **If the MSI cluster genes (Step 4 above) are disproportionately
  flagged relative to other genes**: this is the expected signature of
  the known correlation structure slowing mixing for those specific
  genes, not a general sampler problem -- consider running a longer
  chain (`mcmc_n_iter`) specifically to improve ESS for that cluster, or
  interpreting their posterior intervals with the correlation in mind
  (as already recommended in the project progress log's multicollinearity
  write-up) rather than over-trusting tight-looking but poorly-mixed
  intervals.
- **If flags are widespread across unrelated genes**: increase
  `mcmc_n_iter` and/or `mcmc_thin` (thinning more reduces autocorrelation
  in the kept samples, at the cost of needing more total iterations for
  the same number of kept samples) and re-run Part 3 Step 4 before trusting
  any posterior summary from this fit.
