---
title: "K-Fold Cross-Validated Predictive Likelihood Comparison"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This notebook computes the genuine out-of-sample $k$-fold cross-validated
predictive likelihood comparison specified in `pathway.tex` Phase 3, item
3, following Samorodnitsky et al. (2020) Section 2.4 -- replacing the
quick in-sample-only comparison used so far (`gibbs_sampler_full.Rmd`
Part 3 Step 6) with the rigorous metric originally planned.

**Why this matters, and why it wasn't done first:** an in-sample
comparison structurally favors whichever model has more flexibility to
fit the training data exactly, regardless of whether that flexibility
generalizes. This was demonstrated directly during the ablation
comparison: KRAS appeared far more often as a "driver" in the
no-horseshoe model (161 of 295 drugs) than in the full model (1 of 295),
which is the expected signature of removing a sparsity-inducing prior,
not evidence the no-horseshoe model found 160 additional genuine
discoveries. A proper held-out comparison is the only way to check
whether the full model's added structure (network prior, horseshoe)
actually improves out-of-sample prediction, not just in-sample fit.

**Design notes -- read before running:**

- **$k=3$ folds, not the more conventional $k=5$ or $k=10$.** Each fold
  requires fitting all three models (full, no-network, no-horseshoe)
  once, so total model fits = $3k$. At $k=5$ or $k=10$, even at reduced
  chain settings, this becomes a multi-day commitment (estimated 24--48
  hours at the ablation-comparison chain length); $k=3$ keeps this to a
  single extended run (estimated below) while still giving each fold a
  meaningfully large training set ($\sim$634 of 951 cell lines) and test
  set ($\sim$317 cell lines).
- **Folds are constructed at the cell-line level**, identically across
  all three models and re-used for every model, so the comparison is
  apples-to-apples: the same cell lines are held out for the same fold
  regardless of which model is being scored.
- **The comparison metric is the posterior predictive log-likelihood**,
  computed correctly as $\log\left(\frac{1}{S}\sum_{s=1}^S
  p(y_{\text{test}} \mid \theta^{(s)})\right)$ (averaging the
  *likelihood*, not the log-likelihood, across posterior samples $s$,
  via the log-sum-exp trick for numerical stability) -- **not**
  $\frac{1}{S}\sum_s \log p(y_{\text{test}} \mid \theta^{(s)})$, which is
  a different, smaller quantity by Jensen's inequality and was verified
  to differ meaningfully (by 1--2% in early testing) from the correct
  value during development. Getting this distinction right matters,
  since silently using the wrong one would systematically bias every
  model's score in the same direction without affecting the qualitative
  ranking much -- but a published number should be the right one
  regardless.
- **Each posterior sample's own $\sigma_d^2$ is used**, not the
  posterior mean of $\sigma_d^2$, since a proper posterior predictive
  density integrates over the full posterior, not a plug-in point
  estimate.

## Step 1: Set paths and load data

```{r config}
mutation_matrix_path <- "mutation_matrix.csv"
response_matrix_path <- "response_matrix.csv"
network_matrix_path  <- "gene_adjacency_W.csv"
sampler_path         <- "gibbs_sampler.R"
```

```{r load}
source(sampler_path)

mut <- read.csv(mutation_matrix_path, row.names = 1, check.names = FALSE)
resp <- read.csv(response_matrix_path)
W <- as.matrix(read.csv(network_matrix_path, row.names = 1, check.names = FALSE))

model_ids <- colnames(mut)
M_full <- t(as.matrix(mut))
rownames(M_full) <- model_ids
n_models <- length(model_ids)

all_drugs <- sort(unique(resp$drug_id))
cat(sprintf("Loaded: %d genes x %d models, %d drugs, network %d x %d\n",
            nrow(mut), n_models, length(all_drugs), nrow(W), ncol(W)))
```

## Step 2: Build the response matrix and the K folds

```{r build-y-and-folds}
y_full <- matrix(NA, nrow = n_models, ncol = length(all_drugs),
                  dimnames = list(model_ids, as.character(all_drugs)))
for (i in seq_len(nrow(resp))) {
  y_full[as.character(resp$model_id[i]), as.character(resp$drug_id[i])] <- resp$y[i]
}

set.seed(2024)
k_folds <- 3
fold_assignment <- sample(rep(1:k_folds, length.out = n_models))
cat("Fold sizes:", table(fold_assignment), "\n")
cat("(Same fold assignment is reused for every model below, so the comparison is apples-to-apples.)\n")
```

## Step 3: Posterior predictive log-likelihood function

```{r log-lik-function}
log_mean_exp <- function(x) {
  # Numerically stable log(mean(exp(x))), used to correctly average a
  # LIKELIHOOD (not a log-likelihood) across posterior samples -- see
  # the Overview above for why this is the statistically correct
  # quantity, verified to differ from a naive mean(log-lik) during
  # development.
  m <- max(x)
  m + log(mean(exp(x - m))) - log(length(x))
}

compute_held_out_loglik <- function(fit_chains, y_full, test_idx, drug_ids) {
  # Pools posterior samples across all chains in fit_chains, computing
  # the log predictive density for each sample using THAT sample's own
  # sigma2 (not the posterior mean), then combines via log-mean-exp.
  all_ll <- c()
  for (fit in fit_chains) {
    n_kept <- dim(fit$beta_tilde)[3]
    for (s in seq_len(n_kept)) {
      ll <- 0
      for (d in seq_along(drug_ids)) {
        obs_test <- test_idx[!is.na(y_full[test_idx, d])]
        if (length(obs_test) == 0) next
        mu_pred <- fit$alpha[1, d, s] +
          as.numeric(M_full[obs_test, , drop = FALSE] %*% fit$beta_tilde[, d, s])
        ll <- ll + sum(dnorm(y_full[obs_test, d], mu_pred, sqrt(fit$sigma2[d, s]), log = TRUE))
      }
      all_ll <- c(all_ll, ll)
    }
  }
  log_mean_exp(all_ll)
}
```

## Step 4: Settings

Edit these if you want a different chain length, drug subset, or number
of folds. Defaults were chosen to keep total runtime to roughly a day
(see the printed estimate below) while still using the complete drug
panel.

```{r kfold-settings}
n_iter <- 800
burn_in <- 200
thin <- 2
n_chains <- 2
drugs_to_use <- all_drugs  # full 295-drug panel; subset here if needed, e.g. all_drugs[1:50]

n_fits <- k_folds * 3
cat(sprintf("Total model fits required: %d folds x 3 models = %d\n", k_folds, n_fits))
cat("Benchmark this against a single full-scale fit's measured runtime\n")
cat("(see gibbs_sampler_full.Rmd's performance notes) before committing --\n")
cat("at ~45-50 min per chain at n_iter=1500/295 drugs, n_iter=800 here\n")
cat("should run roughly proportionally faster per fit, but multiply by\n")
cat(sprintf("%d total fits x %d chains before estimating total wall time.\n", n_fits, n_chains))
```

## Step 5: Run all folds for all three models

**This is the long-running step.** Consider running this chunk on its
own (not via "Run All") and saving results incrementally, since a
failure partway through should not require restarting from fold 1.

```{r run-kfold, eval=FALSE}
results <- data.frame(model = character(0), fold = integer(0), held_out_loglik = numeric(0))

model_configs <- list(
  full = list(use_network_layer = TRUE, use_horseshoe_layer = TRUE),
  no_network = list(use_network_layer = FALSE, use_horseshoe_layer = TRUE),
  no_horseshoe = list(use_network_layer = TRUE, use_horseshoe_layer = FALSE)
)

y_sub <- y_full[, as.character(drugs_to_use), drop = FALSE]

for (model_name in names(model_configs)) {
  cfg <- model_configs[[model_name]]
  for (fold_i in seq_len(k_folds)) {
    cat(sprintf("\n=== Model: %s | Fold %d/%d ===\n", model_name, fold_i, k_folds))
    test_idx <- which(fold_assignment == fold_i)

    y_train <- y_sub
    y_train[test_idx, ] <- NA

    t0 <- Sys.time()
    fit_chains <- vector("list", n_chains)
    for (chain_i in seq_len(n_chains)) {
      fit_chains[[chain_i]] <- gibbs_sampler(
        y = y_train, M = M_full, group = rep(1, n_models), W = W,
        delta = 0.8, use_group_layer = FALSE,
        use_network_layer = cfg$use_network_layer,
        use_horseshoe_layer = cfg$use_horseshoe_layer,
        n_iter = n_iter, burn_in = burn_in, thin = thin,
        verbose = 0, seed = chain_i
      )
    }
    elapsed <- as.numeric(difftime(Sys.time(), t0, units = "mins"))

    held_out_ll <- compute_held_out_loglik(fit_chains, y_sub, test_idx, drugs_to_use)
    cat(sprintf("  Held-out log-lik: %.2f  (elapsed: %.1f min)\n", held_out_ll, elapsed))

    results <- rbind(results, data.frame(model = model_name, fold = fold_i,
                                           held_out_loglik = held_out_ll))

    # Save incrementally after every fold, so a failure partway through
    # does not lose completed folds.
    write.csv(results, "kfold_results_partial.csv", row.names = FALSE)
  }
}

write.csv(results, "kfold_results.csv", row.names = FALSE)
cat("\nSaved kfold_results.csv\n")
```

```{r reload-if-resuming, eval=FALSE}
# If resuming after an interruption, reload whatever completed so far:
results <- read.csv("kfold_results_partial.csv")
```

## Step 6: Summarize

```{r summarize, eval=FALSE}
summary_df <- aggregate(held_out_loglik ~ model, data = results,
                         FUN = function(x) c(mean = mean(x), sd = sd(x), sum = sum(x)))
summary_df <- do.call(data.frame, summary_df)
names(summary_df) <- c("model", "mean_loglik_per_fold", "sd_loglik_per_fold", "total_loglik")
cat("=== K-fold cross-validated held-out log-likelihood, by model ===\n")
cat("(Higher / less negative is better -- the model assigns higher predictive\n")
cat("density to data it did not see during fitting.)\n\n")
summary_df[order(-summary_df$total_loglik), ]
```

```{r per-fold-detail, eval=FALSE}
cat("=== Per-fold detail ===\n")
results[order(results$model, results$fold), ]
```
