---
title: "Simulation Study Runner: Revised Network-Structured Scenario 2"
output: html_document
---

# Purpose

This R Markdown file contains the simulation-generation and simulation-fitting framework for the no-group-layer simulation study. The main correction is in **Scenario 2**, which is revised to be a cleaner network-prior-favored scenario. Instead of activating only a very small number of connected genes, Scenario 2 now generates a larger network-connected module per drug, with smooth pathway-level effects over the selected module.

The intended workflow is:

1. Generate simulation files for one scenario at a time.
2. Run the Gibbs sampler for one scenario at a time.
3. Save standard metrics and pair-level score files.
4. Use the pair-level files later for sensitivity at matched FDR, matched number selected, and precision--recall analyses if desired.

```{r setup, message=FALSE, warning=FALSE}
suppressWarnings(suppressMessages({
  library(coda)
}))
```

# 1. Required Sampler

Paste your already-tested `gibbs_sampler()` function above this section, or save it as `gibbs_sampler.R` and uncomment the line below.

```{r sampler-source, eval=FALSE}
source("gibbs_sampler.R")
```

```{r rinvgamma}
rinvgamma <- function(n, shape, rate) {
  1 / rgamma(n, shape = shape, rate = rate)
}
```

# 2. Simulation Settings

```{r simulation-settings}

# Reduced-but-structured simulation dimensions
N <- 400
G <- 150
D <- 50
G_net <- 100
K <- 10
n_reps <- 100

delta <- 0.8
sigma_y <- 1.0
alpha_sd <- 0.5

p_min <- 0.01
p_max <- 0.25

# Scenario 1 and Scenario 3 sparse signal settings
n_signal_per_drug <- 5
n_weak_per_drug <- 25

strong_mean <- 1.0
strong_sd <- 0.25
weak_sd <- 0.10

# Revised Scenario 2 setting:
# Use a larger connected pathway module so the network prior has a fair chance to help.
n_network_signal_per_drug <- 20
network_effect_size <- 0.60
network_effect_sd <- 0.15

# MCMC settings
mcmc_n_iter <- 4000
mcmc_burn_in <- 1000
mcmc_thin <- 3
mcmc_n_chains <- 2

samples_per_chain <- (mcmc_n_iter - mcmc_burn_in) / mcmc_thin
cat("Samples kept per chain:", samples_per_chain, "\n")
cat("Total chains:", mcmc_n_chains, "\n")
```

# 3. Simulation Helper Functions

```{r simulation-helper-functions}
make_mock_network <- function(G_net, n_modules = 8, edge_prob_within = 0.20) {
  W <- matrix(0, G_net, G_net)
  modules <- split(seq_len(G_net), cut(seq_len(G_net), n_modules, labels = FALSE))

  for (mod in modules) {
    if (length(mod) > 1) {
      for (i in mod) {
        for (j in mod) {
          if (i < j && runif(1) < edge_prob_within) {
            W[i, j] <- 1
            W[j, i] <- 1
          }
        }
      }
    }
  }

  deg <- rowSums(W)
  isolated <- which(deg == 0)
  for (g in isolated) {
    h <- sample(setdiff(seq_len(G_net), g), 1)
    W[g, h] <- 1
    W[h, g] <- 1
  }

  diag(W) <- 0
  W
}

make_precision <- function(W, delta = 0.8, ridge = 1e-6) {
  D_W <- diag(rowSums(W))
  Q <- D_W - delta * W
  Q <- Q + diag(ridge, nrow(Q))
  Q
}

rmvnorm_chol <- function(mu, Sigma) {
  R <- chol(Sigma)
  as.numeric(mu + t(R) %*% rnorm(length(mu)))
}

simulate_mutations <- function(N, G, p_min = 0.01, p_max = 0.25, p_emp = NULL) {
  if (is.null(p_emp)) {
    p_g <- runif(G, p_min, p_max)
  } else {
    p_g <- p_emp
  }

  M <- matrix(0, N, G)
  for (g in seq_len(G)) {
    M[, g] <- rbinom(N, size = 1, prob = p_g[g])
  }

  colnames(M) <- paste0("Gene", seq_len(G))
  rownames(M) <- paste0("CL", seq_len(N))

  list(M = M, p_g = p_g)
}

simulate_groups <- function(N, K) {
  sample(seq_len(K), size = N, replace = TRUE)
}

simulate_response_nogroup <- function(M_full, beta_true, alpha, sigma_y = 1.0) {
  N <- nrow(M_full)
  D <- ncol(beta_true)
  y_full <- matrix(NA, N, D)

  for (d in seq_len(D)) {
    mu <- alpha[d] + M_full %*% beta_true[, d]
    y_full[, d] <- as.numeric(mu + rnorm(N, 0, sigma_y))
  }

  rownames(y_full) <- rownames(M_full)
  colnames(y_full) <- paste0("Drug", seq_len(D))
  y_full
}

simulate_response_group <- function(M_full, beta_group, alpha_group, group, sigma_y = 1.0) {
  N <- nrow(M_full)
  D <- dim(beta_group)[2]
  y_full <- matrix(NA, N, D)

  for (i in seq_len(N)) {
    k <- group[i]
    for (d in seq_len(D)) {
      mu_id <- alpha_group[k, d] + sum(M_full[i, ] * beta_group[, d, k])
      y_full[i, d] <- mu_id + rnorm(1, 0, sigma_y)
    }
  }

  rownames(y_full) <- rownames(M_full)
  colnames(y_full) <- paste0("Drug", seq_len(D))
  y_full
}

select_network_module <- function(W, n_genes) {
  G_net_local <- nrow(W)
  seed_gene <- sample(seq_len(G_net_local), 1)
  selected <- seed_gene

  while (length(selected) < n_genes) {
    neighbors <- which(rowSums(W[selected, , drop = FALSE]) > 0)
    candidates <- setdiff(neighbors, selected)

    if (length(candidates) == 0) {
      candidates <- setdiff(seq_len(G_net_local), selected)
    }

    selected <- c(selected, sample(candidates, 1))
  }

  selected
}

W <- make_mock_network(G_net = G_net)
network_gene_idx <- seq_len(G_net)
nonnetwork_gene_idx <- setdiff(seq_len(G), network_gene_idx)
```

# 4. Simulation Scenarios

## Scenario 1: Sparse Independent Effects

```{r scenario-1}
simulate_scenario_1 <- function(rep_id) {
  mut <- simulate_mutations(N, G, p_min, p_max)
  M_full <- mut$M

  beta_true <- matrix(0, G, D)
  true_assoc <- matrix(FALSE, G, D)

  for (d in seq_len(D)) {
    active_genes <- sample(seq_len(G), n_signal_per_drug)
    beta_true[active_genes, d] <- rnorm(n_signal_per_drug, strong_mean, strong_sd) *
      sample(c(-1, 1), n_signal_per_drug, replace = TRUE)
    true_assoc[active_genes, d] <- TRUE
  }

  alpha <- rnorm(D, 0, alpha_sd)
  y_full <- simulate_response_nogroup(M_full, beta_true, alpha, sigma_y)

  W_named <- W
  rownames(W_named) <- colnames(M_full)[seq_len(nrow(W_named))]
  colnames(W_named) <- colnames(M_full)[seq_len(ncol(W_named))]

  list(
    scenario = "scenario_1_sparse_independent",
    y_full = y_full,
    M_full = M_full,
    W = W_named,
    group = rep(1, N),
    drugs_to_use = colnames(y_full),
    beta_true = beta_true,
    true_assoc = true_assoc,
    alpha_true = alpha,
    sigma_y = sigma_y,
    network_gene_idx = network_gene_idx,
    nonnetwork_gene_idx = nonnetwork_gene_idx,
    delta = delta
  )
}
```

## Scenario 2: Revised Network-Structured Effects

This is the revised clean network-prior-favored scenario. Each drug has a larger connected network module of active genes. Effects are moderate and network-correlated. This is intended to make the data-generating mechanism more aligned with the network prior.

```{r scenario-2-revised}
simulate_scenario_2 <- function(rep_id,
                                effect_size = network_effect_size,
                                effect_sd = network_effect_sd,
                                n_module_genes = n_network_signal_per_drug) {
  mut <- simulate_mutations(N, G, p_min, p_max)
  M_full <- mut$M

  beta_true <- matrix(0, G, D)
  true_assoc <- matrix(FALSE, G, D)

  Q <- make_precision(W, delta = delta)
  Sigma_net <- solve(Q)

  for (d in seq_len(D)) {
    active_net_positions <- select_network_module(W, n_module_genes)
    active_genes <- network_gene_idx[active_net_positions]

    sign_d <- sample(c(-1, 1), 1)

    network_noise <- rmvnorm_chol(
      mu = rep(0, G_net),
      Sigma = effect_sd^2 * Sigma_net
    )

    beta_true[active_genes, d] <- sign_d * effect_size +
      network_noise[active_net_positions]

    true_assoc[active_genes, d] <- TRUE
  }

  alpha <- rnorm(D, 0, alpha_sd)
  y_full <- simulate_response_nogroup(M_full, beta_true, alpha, sigma_y)

  W_named <- W
  rownames(W_named) <- colnames(M_full)[seq_len(nrow(W_named))]
  colnames(W_named) <- colnames(M_full)[seq_len(ncol(W_named))]

  list(
    scenario = "scenario_2_network_structured",
    description = "Revised network-smooth module-level effects designed to favor the network prior.",
    y_full = y_full,
    M_full = M_full,
    W = W_named,
    group = rep(1, N),
    drugs_to_use = colnames(y_full),
    beta_true = beta_true,
    true_assoc = true_assoc,
    alpha_true = alpha,
    sigma_y = sigma_y,
    network_gene_idx = network_gene_idx,
    nonnetwork_gene_idx = nonnetwork_gene_idx,
    delta = delta,
    effect_size = effect_size,
    effect_sd = effect_sd,
    n_module_genes = n_module_genes
  )
}
```

## Scenario 3: Dense Weak Noise With Sparse Strong Signals

```{r scenario-3}
simulate_scenario_3 <- function(rep_id) {
  mut <- simulate_mutations(N, G, p_min, p_max)
  M_full <- mut$M

  beta_true <- matrix(0, G, D)
  true_assoc <- matrix(FALSE, G, D)
  weak_assoc <- matrix(FALSE, G, D)

  for (d in seq_len(D)) {
    strong_genes <- sample(seq_len(G), n_signal_per_drug)
    weak_genes <- sample(setdiff(seq_len(G), strong_genes), n_weak_per_drug)

    beta_true[strong_genes, d] <- rnorm(n_signal_per_drug, strong_mean, strong_sd) *
      sample(c(-1, 1), n_signal_per_drug, replace = TRUE)
    beta_true[weak_genes, d] <- rnorm(n_weak_per_drug, 0, weak_sd)

    true_assoc[strong_genes, d] <- TRUE
    weak_assoc[weak_genes, d] <- TRUE
  }

  alpha <- rnorm(D, 0, alpha_sd)
  y_full <- simulate_response_nogroup(M_full, beta_true, alpha, sigma_y)

  W_named <- W
  rownames(W_named) <- colnames(M_full)[seq_len(nrow(W_named))]
  colnames(W_named) <- colnames(M_full)[seq_len(ncol(W_named))]

  list(
    scenario = "scenario_3_dense_weak_noise",
    y_full = y_full,
    M_full = M_full,
    W = W_named,
    group = rep(1, N),
    drugs_to_use = colnames(y_full),
    beta_true = beta_true,
    true_assoc = true_assoc,
    weak_assoc = weak_assoc,
    alpha_true = alpha,
    sigma_y = sigma_y,
    network_gene_idx = network_gene_idx,
    nonnetwork_gene_idx = nonnetwork_gene_idx,
    delta = delta
  )
}
```

## Scenario 4: Group-Specific Effects

```{r scenario-4}
simulate_scenario_4 <- function(rep_id, eta_sd = 0.25) {
  mut <- simulate_mutations(N, G, p_min, p_max)
  M_full <- mut$M
  group <- simulate_groups(N, K)

  beta_mean <- matrix(0, G, D)
  true_assoc <- matrix(FALSE, G, D)

  for (d in seq_len(D)) {
    active_genes <- sample(seq_len(G), n_signal_per_drug)
    beta_mean[active_genes, d] <- rnorm(n_signal_per_drug, strong_mean, strong_sd) *
      sample(c(-1, 1), n_signal_per_drug, replace = TRUE)
    true_assoc[active_genes, d] <- TRUE
  }

  beta_group <- array(0, dim = c(G, D, K))
  for (k in seq_len(K)) {
    beta_group[, , k] <- beta_mean + matrix(rnorm(G * D, 0, eta_sd), G, D) * true_assoc
  }

  alpha_group <- matrix(rnorm(K * D, 0, alpha_sd), K, D)
  y_full <- simulate_response_group(M_full, beta_group, alpha_group, group, sigma_y)

  W_named <- W
  rownames(W_named) <- colnames(M_full)[seq_len(nrow(W_named))]
  colnames(W_named) <- colnames(M_full)[seq_len(ncol(W_named))]

  list(
    scenario = "scenario_4_group_specific",
    y_full = y_full,
    M_full = M_full,
    W = W_named,
    group = group,
    drugs_to_use = colnames(y_full),
    beta_true = beta_mean,
    beta_group_true = beta_group,
    true_assoc = true_assoc,
    alpha_group_true = alpha_group,
    sigma_y = sigma_y,
    eta_sd = eta_sd,
    network_gene_idx = network_gene_idx,
    nonnetwork_gene_idx = nonnetwork_gene_idx,
    delta = delta
  )
}
```

## Scenario 5: Network and Group Heterogeneity

```{r scenario-5}
simulate_scenario_5 <- function(rep_id,
                                effect_size = network_effect_size,
                                effect_sd = network_effect_sd,
                                eta_sd = 0.25,
                                n_module_genes = n_network_signal_per_drug) {
  mut <- simulate_mutations(N, G, p_min, p_max)
  M_full <- mut$M
  group <- simulate_groups(N, K)

  beta_mean <- matrix(0, G, D)
  true_assoc <- matrix(FALSE, G, D)

  Q <- make_precision(W, delta = delta)
  Sigma_net <- solve(Q)

  for (d in seq_len(D)) {
    active_net_positions <- select_network_module(W, n_module_genes)
    active_genes <- network_gene_idx[active_net_positions]

    sign_d <- sample(c(-1, 1), 1)
    network_noise <- rmvnorm_chol(rep(0, G_net), effect_sd^2 * Sigma_net)

    beta_mean[active_genes, d] <- sign_d * effect_size +
      network_noise[active_net_positions]
    true_assoc[active_genes, d] <- TRUE
  }

  beta_group <- array(0, dim = c(G, D, K))
  for (k in seq_len(K)) {
    beta_group[, , k] <- beta_mean + matrix(rnorm(G * D, 0, eta_sd), G, D) * true_assoc
  }

  alpha_group <- matrix(rnorm(K * D, 0, alpha_sd), K, D)
  y_full <- simulate_response_group(M_full, beta_group, alpha_group, group, sigma_y)

  W_named <- W
  rownames(W_named) <- colnames(M_full)[seq_len(nrow(W_named))]
  colnames(W_named) <- colnames(M_full)[seq_len(ncol(W_named))]

  list(
    scenario = "scenario_5_network_group",
    y_full = y_full,
    M_full = M_full,
    W = W_named,
    group = group,
    drugs_to_use = colnames(y_full),
    beta_true = beta_mean,
    beta_group_true = beta_group,
    true_assoc = true_assoc,
    alpha_group_true = alpha_group,
    sigma_y = sigma_y,
    eta_sd = eta_sd,
    effect_size = effect_size,
    effect_sd = effect_sd,
    n_module_genes = n_module_genes,
    network_gene_idx = network_gene_idx,
    nonnetwork_gene_idx = nonnetwork_gene_idx,
    delta = delta
  )
}
```

# 5. Scenario Generator List

```{r scenario-generators}
scenario_generators <- list(
  scenario_1_sparse_independent = simulate_scenario_1,
  scenario_2_network_structured = simulate_scenario_2,
  scenario_3_dense_weak_noise = simulate_scenario_3,
  scenario_4_group_specific = simulate_scenario_4,
  scenario_5_network_group = simulate_scenario_5
)
```

# 6. Generate Simulation Files

```{r generate-one-scenario-function}
generate_one_scenario_files <- function(scenario_name,
                                        out_dir = ".",
                                        n_reps = 100,
                                        overwrite = FALSE) {
  if (!scenario_name %in% names(scenario_generators)) {
    stop("Unknown scenario_name: ", scenario_name)
  }

  scenario_dir <- file.path(out_dir, scenario_name)
  dir.create(scenario_dir, showWarnings = FALSE, recursive = TRUE)

  gen_fun <- scenario_generators[[scenario_name]]

  for (rep_id in seq_len(n_reps)) {
    sim_file <- file.path(
      scenario_dir,
      paste0("sim_", scenario_name, "_rep", sprintf("%03d", rep_id), ".rds")
    )

    if (file.exists(sim_file) && !overwrite) {
      cat("Skipping existing:", sim_file, "\n")
      next
    }

    sim <- gen_fun(rep_id)
    saveRDS(sim, sim_file)
    cat("Saved:", sim_file, "\n")
  }

  invisible(TRUE)
}
```

Generate one scenario at a time. For the revised Scenario 2, use `overwrite = TRUE` only when you intentionally want to replace the old Scenario 2 files.

```{r generate-scenario-example, eval=FALSE}
generate_one_scenario_files(
  scenario_name = "scenario_2_network_structured",
  out_dir = ".",
  n_reps = 100,
  overwrite = TRUE
)
```

# 7. Model Configurations and Metric Functions

```{r model-configs-and-score-functions}
model_configs_nogroup <- list(
  full = list(use_group_layer = FALSE, use_network_layer = TRUE, use_horseshoe_layer = TRUE),
  no_network = list(use_group_layer = FALSE, use_network_layer = FALSE, use_horseshoe_layer = TRUE),
  no_horseshoe = list(use_group_layer = FALSE, use_network_layer = TRUE, use_horseshoe_layer = FALSE)
)

model_configs_group <- list(
  full_group = list(use_group_layer = TRUE, use_network_layer = TRUE, use_horseshoe_layer = TRUE),
  no_network_group = list(use_group_layer = TRUE, use_network_layer = FALSE, use_horseshoe_layer = TRUE),
  no_horseshoe_group = list(use_group_layer = TRUE, use_network_layer = TRUE, use_horseshoe_layer = FALSE),
  full_no_group = list(use_group_layer = FALSE, use_network_layer = TRUE, use_horseshoe_layer = TRUE)
)

combine_beta_draws_base <- function(fit_chains) {
  beta_list <- lapply(fit_chains, function(fit) fit$beta_tilde)
  dims <- lapply(beta_list, dim)
  G_local <- dims[[1]][1]
  D_local <- dims[[1]][2]
  total_samples <- sum(sapply(dims, function(x) x[3]))

  beta_combined <- array(NA, dim = c(G_local, D_local, total_samples))
  start_idx <- 1
  for (i in seq_along(beta_list)) {
    n_i <- dim(beta_list[[i]])[3]
    end_idx <- start_idx + n_i - 1
    beta_combined[, , start_idx:end_idx] <- beta_list[[i]]
    start_idx <- end_idx + 1
  }
  beta_combined
}

make_pair_scores <- function(fit_chains, beta_true, true_assoc,
                             scenario, replicate, replicate_file, model) {
  beta_draws <- combine_beta_draws_base(fit_chains)

  beta_hat <- apply(beta_draws, c(1, 2), mean)
  ci_low <- apply(beta_draws, c(1, 2), quantile, probs = 0.025, na.rm = TRUE)
  ci_high <- apply(beta_draws, c(1, 2), quantile, probs = 0.975, na.rm = TRUE)

  p_pos <- apply(beta_draws, c(1, 2), function(x) mean(x > 0, na.rm = TRUE))
  p_neg <- apply(beta_draws, c(1, 2), function(x) mean(x < 0, na.rm = TRUE))
  sign_prob <- pmax(p_pos, p_neg)

  selected_ci <- (ci_low > 0) | (ci_high < 0)

  data.frame(
    scenario = scenario,
    replicate = replicate,
    replicate_file = replicate_file,
    model = model,
    gene_index = rep(seq_len(nrow(beta_hat)), times = ncol(beta_hat)),
    drug_index = rep(seq_len(ncol(beta_hat)), each = nrow(beta_hat)),
    beta_hat = as.vector(beta_hat),
    abs_beta_hat = abs(as.vector(beta_hat)),
    sign_prob = as.vector(sign_prob),
    ci_low = as.vector(ci_low),
    ci_high = as.vector(ci_high),
    selected_ci = as.vector(selected_ci),
    beta_true = as.vector(beta_true),
    true_assoc = as.vector(true_assoc)
  )
}

evaluate_metrics_from_pair_scores <- function(pair_scores) {
  selected <- as.logical(pair_scores$selected_ci)
  true_assoc <- as.logical(pair_scores$true_assoc)

  TP <- sum(selected & true_assoc, na.rm = TRUE)
  FP <- sum(selected & !true_assoc, na.rm = TRUE)
  FN <- sum(!selected & true_assoc, na.rm = TRUE)
  TN <- sum(!selected & !true_assoc, na.rm = TRUE)

  data.frame(
    rmse_beta = sqrt(mean((pair_scores$beta_hat - pair_scores$beta_true)^2, na.rm = TRUE)),
    sensitivity = ifelse((TP + FN) > 0, TP / (TP + FN), NA),
    specificity = ifelse((TN + FP) > 0, TN / (TN + FP), NA),
    precision = ifelse((TP + FP) > 0, TP / (TP + FP), NA),
    fdr = ifelse((TP + FP) > 0, FP / (TP + FP), NA),
    coverage_95 = mean(pair_scores$beta_true >= pair_scores$ci_low &
                         pair_scores$beta_true <= pair_scores$ci_high, na.rm = TRUE),
    n_selected = sum(selected, na.rm = TRUE),
    true_nonzero = sum(true_assoc, na.rm = TRUE),
    TP = TP,
    FP = FP,
    FN = FN,
    TN = TN
  )
}
```

# 8. Run One Scenario Simulation

```{r run-one-scenario-simulation-fixed}
run_one_scenario_simulation <- function(scenario_name,
                                        base_dir = ".",
                                        results_dir = "simulation_results",
                                        model_configs = model_configs_nogroup,
                                        max_reps = 100,
                                        save_fits = FALSE,
                                        save_pair_scores = TRUE) {
  dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

  scenario_dir <- file.path(base_dir, scenario_name)
  if (!dir.exists(scenario_dir)) stop("Scenario directory does not exist: ", scenario_dir)

  sim_files <- sort(list.files(scenario_dir, pattern = "\\.rds$", full.names = TRUE))
  sim_files <- sim_files[seq_len(min(max_reps, length(sim_files)))]

  partial_file <- file.path(results_dir, paste0("metrics_", scenario_name, "_partial.csv"))
  final_file <- file.path(results_dir, paste0("metrics_", scenario_name, "_final.csv"))

  if (file.exists(partial_file)) {
    results <- read.csv(partial_file)
  } else {
    results <- data.frame()
  }

  for (rep_i in seq_along(sim_files)) {
    sim_file <- sim_files[rep_i]
    sim <- readRDS(sim_file)

    rep_id_text <- tools::file_path_sans_ext(basename(sim_file))
    rep_num <- as.integer(sub(".*_rep([0-9]+)$", "\\1", rep_id_text))

    y_real <- sim$y_full
    M_real <- sim$M_full
    W_real <- sim$W

    rownames(W_real) <- colnames(M_real)[seq_len(nrow(W_real))]
    colnames(W_real) <- colnames(M_real)[seq_len(ncol(W_real))]

    beta_true <- as.matrix(sim$beta_true)
    true_assoc <- as.matrix(sim$true_assoc)

    for (model_name in names(model_configs)) {
      if (nrow(results) > 0) {
        already_done <- any(results$scenario == scenario_name &
                              results$replicate == rep_num &
                              results$model == model_name)
      } else {
        already_done <- FALSE
      }

      if (already_done) {
        cat("Skipping completed:", scenario_name, "rep", rep_num, model_name, "\n")
        next
      }

      cfg <- model_configs[[model_name]]
      group_vec <- if (cfg$use_group_layer) sim$group else rep(1, nrow(y_real))

      cat("\n===", scenario_name, "| replicate", rep_num, "| model", model_name, "===\n")

      fit_chains <- vector("list", mcmc_n_chains)
      t0 <- Sys.time()

      for (chain_i in seq_len(mcmc_n_chains)) {
        fit_chains[[chain_i]] <- gibbs_sampler(
          y = y_real,
          M = M_real,
          group = group_vec,
          W = W_real,
          delta = 0.8,
          use_group_layer = cfg$use_group_layer,
          use_network_layer = cfg$use_network_layer,
          use_horseshoe_layer = cfg$use_horseshoe_layer,
          n_iter = mcmc_n_iter,
          burn_in = mcmc_burn_in,
          thin = mcmc_thin,
          verbose = 200,
          seed = 100000 * rep_num + 1000 * which(names(model_configs) == model_name) + chain_i
        )
      }

      elapsed_mins <- as.numeric(difftime(Sys.time(), t0, units = "mins"))

      pair_scores <- make_pair_scores(
        fit_chains = fit_chains,
        beta_true = beta_true,
        true_assoc = true_assoc,
        scenario = scenario_name,
        replicate = rep_num,
        replicate_file = basename(sim_file),
        model = model_name
      )

      metrics <- evaluate_metrics_from_pair_scores(pair_scores)
      metrics$scenario <- scenario_name
      metrics$replicate <- rep_num
      metrics$replicate_file <- basename(sim_file)
      metrics$model <- model_name
      metrics$elapsed_mins <- elapsed_mins
      metrics$n_iter <- mcmc_n_iter
      metrics$burn_in <- mcmc_burn_in
      metrics$thin <- mcmc_thin
      metrics$n_chains <- mcmc_n_chains

      results <- rbind(results, metrics)
      write.csv(results, partial_file, row.names = FALSE)

      if (save_pair_scores) {
        score_dir <- file.path(results_dir, "pair_scores", scenario_name, model_name)
        dir.create(score_dir, showWarnings = FALSE, recursive = TRUE)
        score_file <- file.path(score_dir, paste0("pair_scores_", scenario_name,
                                                  "_rep", sprintf("%03d", rep_num),
                                                  "_", model_name, ".rds"))
        saveRDS(pair_scores, score_file)
      }

      if (save_fits) {
        fit_dir <- file.path(results_dir, "saved_fits", scenario_name, model_name)
        dir.create(fit_dir, showWarnings = FALSE, recursive = TRUE)
        fit_file <- file.path(fit_dir, paste0("fit_", scenario_name,
                                              "_rep", sprintf("%03d", rep_num),
                                              "_", model_name, ".rds"))
        saveRDS(fit_chains, fit_file)
      }

      rm(fit_chains, pair_scores)
      gc()
    }
  }

  write.csv(results, final_file, row.names = FALSE)
  invisible(results)
}
```

Run one scenario at a time.

```{r run-scenario-example, eval=FALSE}
scenario_results <- run_one_scenario_simulation(
  scenario_name = "scenario_2_network_structured",
  base_dir = ".",
  results_dir = "simulation_results",
  model_configs = model_configs_nogroup,
  max_reps = 100,
  save_fits = FALSE,
  save_pair_scores = TRUE
)
```
